---
title: "Scenarios, custom registers, and bias testing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scenarios, custom registers, and bias testing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This picks up where the README's 5-minute quickstart leaves off. If you
haven't read that yet, start there — it covers install, generating your
first registers, saving to files, and joining tables. Everything below
assumes you already have a `schema` and a `pop`:

```{r setup}
library(fiktive)

schema <- load_registers_schema()
pop <- generate_background_population(n = 200, seed = 1, schema = schema)
```

## Structural noise vs. realistic-looking data

By default, fiktive generates **structural noise that joins** — every code
is valid, but drawn uniformly, with no attempt to look like real Denmark.
A municipality column is exactly as likely to say Læsø (pop. 1,655) as
Copenhagen (pop. 670,389).

Pass `constraints=` to `generate_register()` / `generate_registers()` to
opt into one or more real-world-shaped defaults instead. Two naming
families, by mechanism: `valid_*` never emits a definitionally impossible
combination; `weighted_*` samples a categorical column by its real-world
empirical distribution instead of uniformly.

```{r constraints}
tables <- generate_registers(
  registers = c("bef", "lpr_adm", "lpr_diag"),
  population = pop, schema = schema,
  from = as.Date("2008-01-01"), to = as.Date("2009-12-31"),
  seed = 1,
  constraints = c("valid_diagnosis_sex_age", "weighted_municipality")   # opt-in: none set by default
)
```

Right now three constraints exist: `"weighted_municipality"` (`kom`
weighted by real 2026 population instead of drawn evenly);
`"valid_diagnosis_sex_age"` (diagnosis codes, `icd10`/`icd10_sks`, never
assign a chapter impossible for the patient's sex or age -- no pregnancy
code on a man, no perinatal code on someone past infancy); and
`"weighted_lprindberetningssystem"` (LPR3's `lprindberetningssystem` --
which historical reporting system a contact came through: `LPR3`, `LPR2`,
`MiniPAS`, or `LPR1` -- weighted toward `LPR3` instead of drawn evenly
across all four). An unknown name in `constraints=` errors rather than
being silently ignored.

`kont_type`'s own format (an SKS admin code for `LPR3` rows, a legacy LPR2
patient-type digit for the other three) always follows whatever
`lprindberetningssystem` that row got, regardless of which constraints are
set -- two columns on the same row have to agree with each other no matter
what, so this isn't something to opt into separately.

This is about **plausibility, not planted signal** — it never changes what
any `scenario` claims or what `get_truth()` reports. If you want a
specific, known, testable relationship instead of realistic-looking
background shape, that's what `scenario_association()` and friends (below)
are for.

## What do these columns and codes actually mean?

Register columns often have short, cryptic names (`civst`, `hfaudd`,
`fm_mark`), and their coded values are worse (`civst = "U"`, `koen = 1`).
`codebook()` looks up the real description for each, straight from the
same schema that generated your data — in Danish and English:

```{r codebook}
codebook(schema, "bef")
```

Pass more than one id (`codebook(schema, c("bef", "lmdb"))`) to get several
registers' columns at once, with a `register` column added so you can tell
them apart. `NA` in `label_da`/`label_en`/`values` just means the schema
doesn't document one for that column — same "don't invent it" rule as
everywhere else in fiktive.

## Describing your own register (e.g. a study cohort)

Sometimes you have a table that isn't a real DST register — a study
cohort, a set of questionnaire scores, a group label. `generate_custom_register()`
makes one up structurally (you describe the columns; fiktive fills in
plausible fake values) and it joins to everything else via `pnr`:

```{r custom-register}
# Describe your columns: a type, plus either a min/max range or a set of values
cols <- tibble::tibble(
  name   = c("score", "grp"),
  type   = c("integer", "character"),
  min    = c(0, NA),
  max    = c(10, NA),
  values = c(NA, "A|B|C")
)

ext <- generate_custom_register(
  id = "my_study",   # any name you want for this table
  one_row_per = "person_reference_date",   # one row per person per snapshot date
  columns = cols,
  population = pop,
  schema = schema,
  from = as.Date("2008-01-01"),   # dates are just an example, use your own window
  to = as.Date("2009-12-31"),
  seed = 1,
  cadence = "annual"   # one snapshot per year ("quarterly" is the other option)
)

joined <- dplyr::inner_join(tables$bef, ext, by = "pnr", relationship = "many-to-many")
```

### Describing many columns with a CSV instead

For a real cohort with dozens or hundreds of variables, typing out a
tibble by hand doesn't scale. Pass a CSV path instead of a tibble — same
`name` / `type` / `min` / `max` / `values` columns, one row per variable.
Any other columns already in your file (a label, a request status, a
source note — whatever you're tracking for your own bookkeeping) are
simply ignored, so your documentation columns can live right next to the
ones fiktive reads:

```csv
name,type,min,max,values,dataset,label,status,source
id,integer,1,60000,,journal,KKH ID number,available,catalogue
mdato,date,1993-12-01,1997-05-31,,journal,Date of participation,requested,catalogue
center,character,,,KBH|AAR,journal,Study center,requested,catalogue
kqn,character,,,M|K,journal,Gender,requested,catalogue
vaegt,numeric,42,145,,journal,Weight (kg),requested,catalogue
fedtbiop,integer,,,0|1,journal,Fat biopsy taken (yes/no),available,catalogue
```

```{r custom-register-csv, eval = FALSE}
ext <- generate_custom_register(
  id = "kkh",
  one_row_per = "person_reference_date",
  columns = "kkh_columns.csv",   # path to the CSV above
  population = pop,
  schema = schema,
  from = as.Date("1993-12-01"),
  to = as.Date("1997-05-31"),
  seed = 1,
  cadence = "annual"
)
```

`values` accepts `|` or `;` as the separator (`KBH|AAR`, `0|1`, ...).
Rows with neither `min`/`max` nor `values` filled in still work — fiktive
falls back to a generic default for that type, so a work-in-progress CSV
where you haven't decided every range yet won't error out.

## Checking whether your analysis code is actually correct

Normally when you test analysis code, you don't actually know what the
right answer is supposed to be — you're just checking that it runs and the
output looks plausible. fiktive lets you flip that around: you tell it a
real, exact relationship to secretly build into the data ("increasing `x`
by 1 always increases `y` by 1.5, on average"), it generates fake data with
that relationship baked in, and it hands you back that true number. You
then run **your own** analysis code on the fake data and check whether it
finds the same number. If it doesn't, the bug is in your code — not the
data.

Three steps: **describe the relationship to plant → generate data with
it → compare your analysis against fiktive's true answer.**

**Step 1 — describe the relationship.** `scenario_association()` says
"column `x` affects column `y`, and here's the true effect size":

```{r association-scenario}
# "increasing x by 1 increases y by 1.5, on average" -- the true, known answer
sc <- scenario_association(exposure = "study.x", outcome = "study.y", coefficient = 1.5)
```

`exposure` is the column doing the affecting, `outcome` is the column
being affected, `coefficient` is the true effect size (change this to
whatever number you want to test against). `"study.x"` means "column `x`
on the table called `study`" — the table you generate next.

**Step 2 — generate the data with that relationship.** Same
`generate_custom_register()` as above, just with `scenario = sc` added:

```{r association-generate}
cols <- tibble::tibble(
  name = c("x", "y"), type = c("numeric", "numeric"),
  min = c(-2, -2), max = c(2, 2)
)

study <- generate_custom_register(
  id = "study", one_row_per = "person_reference_date", columns = cols,
  population = pop, schema = schema,
  from = as.Date("2008-01-01"), to = as.Date("2008-12-31"),
  seed = 1, scenario = sc, cadence = "annual"
)
```

**Step 3 — run your own analysis, then compare it to the true answer.**
`get_truth()` gives you back the answer fiktive planted — it comes free
with anything you generate, you never pass it in yourself:

```{r association-check}
fit <- lm(y ~ x, data = study)
coef(fit)[["x"]]                  # your analysis's answer -- should come out close to 1.5

get_truth(study)$expected_naive   # fiktive's true answer: 1.5
```

If those two numbers are close, your analysis code works. If they're way
off, something in your code needs fixing — not the data.

Beyond a plain association, fiktive can plant trickier situations, so you
can test whether your code handles them correctly too:

| Function | Plants... | Use it to test... |
|---|---|---|
| `scenario_association()` | a straight X → Y effect | your basic model recovers the right coefficient |
| `scenario_confounding()` | a third variable biasing the naive estimate | your code actually adjusts for confounders |
| `scenario_mnar()` | missing values that depend on the value itself | your code doesn't ignore informative missingness |
| `scenario_complete_case()` | rows dropped depending on a column's value | your "complete case" analysis isn't secretly biased |
| `scenario_misclassification()` | a coded column's values swapped for other real codes | your code isn't thrown off by mislabeled diagnosis/drug codes |

All five work the same way as the example above: build the scenario, pass
it as `scenario=`, then compare your own fit against
`get_truth(x)$expected_naive` (the answer a straightforward analysis
should find) and `$expected_adjusted` (the answer after doing it
properly — e.g. adjusting for the confounder). For a plain association
those two are the same number; for the trickier scenarios they're
deliberately different, which is exactly what lets you test whether your
code does the adjustment correctly. Run `?scenario_confounding`,
`?scenario_mnar`, `?scenario_complete_case`, or `?scenario_misclassification`
for each one's full parameter list and a runnable example.

Leave `scenario=` out entirely (or generate any ordinary register like
`bef`) and you get **independence** — no planted relationship;
`get_truth(x)$expected_naive` comes back `0`.

## Testing for immortal time bias

This one works differently from the scenarios above: instead of adding a
bias on top of an ordinary table, it generates a **survival cohort** —
people with an entry time, an exposure that starts at some point during
follow-up (or never), and an event (or censoring) time. Immortal time bias
is a classic mistake: treating "ever exposed" as if it were true from the
start, which credits exposed people with survival time before their
exposure actually began — making the exposure look protective even when it
does nothing.

```{r immortal-time}
sc <- scenario_immortal_time(
  baseline_hazard = 0.1,     # unexposed event rate per year
  true_hazard_ratio = 1,     # 1 = exposure truly does nothing -- isolates the bias
  exposure_rate = 0.2,       # how quickly people who ever get exposed, do
  horizon_years = 5          # match this to your to - from window below
)

cohort <- generate_custom_register(
  id = "cohort",
  one_row_per = "time_to_event",   # fixed shape: entry_time, exit_time, event, exposure_start_time, ever_exposed
  population = pop,
  schema = schema,
  from = as.Date("2010-01-01"),
  to = as.Date("2015-01-01"),      # 5 years, matching horizon_years above
  seed = 1,
  scenario = sc
)

# The mistake: exposure treated as fixed from the start
fit <- survival::coxph(survival::Surv(entry_time, exit_time, event) ~ ever_exposed, data = cohort)
exp(coef(fit)[["ever_exposed"]])  # hazard ratio well below 1, even though true_hazard_ratio = 1

get_truth(cohort)$expected_naive     # the biased answer the mistake above produces
get_truth(cohort)$expected_adjusted  # 1 -- the true answer, recovered by correctly
                                      # treating exposure as time-varying (see ?scenario_immortal_time
                                      # for a worked example using survival::tmerge())
```

`one_row_per = "time_to_event"` always needs a `scenario_immortal_time()`
or `scenario_left_truncation()` (below) — there's no independence version
of this grain, and `columns=`/`fidelity=` don't apply to it (its shape is
fixed, not user-described).

## Testing for left truncation bias

Also a `time_to_event` register, but a genuinely different mistake from
immortal time: this one is about **delayed entry**. People only enter a
cohort at some age, not at birth, and are only observed at all if they
survived to that age. Analyzing "time since entry" instead of properly
accounting for age at entry biases the result whenever risk actually
depends on age (which it usually does — older people are usually at
higher risk of most things):

```{r left-truncation}
sc <- scenario_left_truncation(
  shape = 5,                # how strongly risk increases with age (1 = no age effect, no bias to demonstrate)
  scale = 80,                # a characteristic age for the outcome, e.g. lifetime/mortality-style
  true_hazard_ratio = 1.5,   # true effect of `group` -- must not be 1, or there's nothing to bias
  max_entry_age = 70         # people enter at a random age between 0 and this
)

cohort <- generate_custom_register(
  id = "cohort",
  one_row_per = "time_to_event",   # fixed shape: entry_age, exit_age, event, group
  population = pop,
  schema = schema,
  from = as.Date("2010-01-01"),    # this scenario's clock is age, not calendar time --
  to = as.Date("2015-01-01"),      # from/to don't affect it, just needed by generate_custom_register()
  seed = 1,
  scenario = sc
)

# The mistake: age at entry discarded, "time since entry" used instead
fit <- survival::coxph(survival::Surv(exit_age - entry_age, event) ~ group, data = cohort)
exp(coef(fit)[["group"]])  # biased toward 1, even though true_hazard_ratio = 1.5

get_truth(cohort)$expected_naive     # the biased answer the mistake above produces
get_truth(cohort)$expected_adjusted  # 1.5 -- the true answer, recovered by correctly
                                      # declaring entry_age as the left-truncation point:
                                      # survival::coxph(survival::Surv(entry_age, exit_age, event) ~ group, data = cohort)
```

## Making data messier (to test your pipeline's robustness)

Real data has missing values and the occasional wild outlier. To rehearse
how your code handles that, ask for `fidelity = "messy"` on any generate
call:

```{r messy}
bef_messy <- generate_register(
  "bef", pop, schema,
  from = as.Date("2008-01-01"), to = as.Date("2009-12-31"),
  seed = 1,
  fidelity = "messy"   # sprinkles in some NAs and extreme values
)
```

Default is `fidelity = "clean"` (no artificial noise) — use that whenever
you're checking `get_truth()` against a scenario, so cosmetic messiness
doesn't get mixed up with the relationship you actually planted.
